Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-24 model.tex
reduced red font in related

      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colLMAForecast.png}
    \caption{\col \\ LMA}
    \label{fig:colLMA}
  \end{subfigure}
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccLMAForecast.png}
    \caption{\gcc \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}  
        \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast.png}
    \caption{\svdfive \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   \caption{Predicted ($p_j$) versus true values ($c_j$) for
     forecasts of \col, \gcc, and \svdfive.  On this type of plot, a
     perfect prediction lies exactly on the diagonal.
%
{\color{red} Joshua, what would you think of adding a row of
  random-walk plots at the top?  Also, the letters in the
  subcaptions---(a) (b) etc.---feel clunky here.  What would you think
  of removing them?}
%
}
\label{fig:forecast-example}
  %\end{subfigure}
\end{figure*} 
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while ARIMA
does not.  {\color{red}[[We should comment on what a noise ball means
      on such a plot, as well as what a horizontal line means.]]}

>>>>

      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colLMAForecast.png}
    \caption{\col \\ LMA}
    \label{fig:colLMA}
  \end{subfigure}
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccLMAForecast.png}
    \caption{\gcc \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}  
        \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast.png}
    \caption{\svdfive \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   \caption{Predicted ($p_j$) versus true values ($c_j$) for
     forecasts of \col, \gcc, and \svdfive.  On this type of plot, a
     perfect prediction lies exactly on the diagonal.
%
{\color{red} Joshua, what would you think of adding a row of
  random-walk plots at the top?  Also, the letters in the
  subcaptions---(a) (b) etc.---feel clunky here.  What would you think
  of removing them?}
%
}
\label{fig:forecast-example}
  %\end{subfigure}
\end{figure*} 
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while ARIMA
does not.  {\color{red}[[We should comment on what a noise ball means
      on such a plot, as well as what a horizontal line means. maybe
      also tie back to the comment on
      page~\pageref{page:normal-errors}.]]}

### 2014-03-24 related.tex
reduced red font in related

Quantifying predictability, which is sometimes called ``predicting
predictability,'' is not a new problem.  Most solutions to it fall
into two categories: model-based error analysis and model-free
information analysis.
%
%Actually this first class are both just analyzing model error distributions but some are local and some are global models but both are really just error
%
%predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
%
The first class focuses on errors produced by a fixed forecasting
schema.  This analysis can proceed locally or globally.  The local
version approximates error distributions for different regions of a
time-series model using local ensemble in-sample forecasting.
{\color{red}[[Joshua, we probably want a quick definition of in-sample
      and out-of-sample.]]} These distributions are then used to
estimate out-of-sample forecast errors {\color{red}[[in that area of]]
  [[means?]]} the time series model.  For example, Smith {\sl et al.}
make in-sample forecasts using ensembles around selected points in
order to predict the local predictability of that time
series~\cite{Smith199250}.  This approach can be used to show that
different portions of a time series can exhibit varying levels of
local predictive uncertainty.  We expand on this finding later in this
paper with a time series that exhibits interesting regime shifts.

Local model-based error analysis works quite well, but it only
approximates the \emph{local} predictive uncertainty \emph{in relation
  to a fixed model}.  It cannot quantify the inherent predictability
of a time series and thus cannot be used to draw conclusions about
predictive structure that may be usable by other forecast methods.
%
%Distrubtion of error. For many methods, if your error is not normally distributed this signals that there is stil more predictve structure to that could be used by for example a larger order ARMA process. But if error is normally distributed, this suggests that you have used up all the predictive structure that **that** model can use, this doesn't quantify if preidctuce structure exists that isn't being used by this process. For example, nonlinear structure which is ignored by a linear predictor.
%
Global model-based error analysis moves in this direction.  It uses
out-of-sample error distributions, computed \emph{post facto} from a
class of models, to determine which of those models was best.  After
building an autoregressive model, for example, it is common to
calculate the errors and verify that they are normally distributed.
If they are not, that indicates that there is structure in the time
series that the model-building process was unable to capture and use.
The problem with this approach is lack of generality.  Normally
distributed errors indicate that a model has captured the structure in
the data insofar as is possible, \emph{given the formulation of that
  particular model} (viz., the best possible linear fit to a nonlinear
dataset).  This gives no indication as to whether another modeling
strategy might do better.  {\color{red} [[I'm not sure I believe that
      a model's bad fit wouldn't manifest in the error distribution.
      Is that really true?]]}

A practice known as deterministic/sto\-chas\-tic
modeling~\cite{Casdagli92dvsplots, weigend-book} bridges the gap
between local and global model-based error analysis {\color{red} [[or
      also in model-free?]]}  The basic idea is to construct a series
of local linear fits, beginning with a few points and working up to to
a global linear fit that includes all known points, and then analyze
how the average out-of-sample error changes as a function of number of
points in the fit. The shape of such a graph indicates the amounts of
determinism and stochasticity are present in a time series.

@@

We build on the notion of \emph{redundancy} that was introduced on
page~\pageref{page:redundancy}, which formally quantifies how
information propagates forward through a time series:
% how much information prior observations of a system lend to future
% values
e.g., the mutual information between the past $n$ observations and the
current one.  {\color{red} [[How is redundancy different from/more
      general than MI, then?]]}  For example, this would be the amount
of information that delay coordinate embedding captures.  {\color{red}
  [[This will be very hard to understand, coming as it does before the
      description of that method.  Does it add something essential
      here?]]}  The redundancy of i.i.d. random processes is zero,
since all observations are independent of one another.  On the other
hand, any deterministic system --- including chaotic ones --- can be
perfectly predicted if observed for long enough~\cite{weigend-book}.
{\color{red} [[Previous sentence leaves the connection to redundancy
      implicit.  Need to make it explicit.]]}  
In practice, it is quite difficult to estimate the redundancy of an
arbitrary, real-valued time series.  Doing so requires knowing either
the Kolmolgorov-Sinai entropy or the values of all positive Lyapunov
exponents of the system.  Both of these calculations are
difficult---the latter particularly so if the data are very noisy or
the generating system is stochastic.

>>>>

Quantifying predictability, which is sometimes called ``predicting
predictability,'' is not a new problem.  Most solutions to it fall
into two categories: model-based error analysis and model-free
information analysis.
%
%Actually this first class are both just analyzing model error distributions but some are local and some are global models but both are really just error
%
%predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
%
The first class focuses on errors produced by a fixed forecasting
schema.  This analysis can proceed locally or globally.  The local
version approximates error distributions for different regions of a
time-series model using local ensemble in-sample
forecasting\footnote{[[The terms ``in sample'' and ``out of sample''
      are used in different ways in the forecasting community.  By in
      sample, we mean a ``forecast'' that is done on the stuff that
      you know, for the purposes of assessing models, errors, and
      prediction horizons.  An out-of-sample prediction is what one
      thinks of as a prediction: forecasting the future.]]}.  These
distributions are then used as estimates of out-of-sample forecast
errors in those regions.  For example, Smith {\sl et al.}  make
in-sample forecasts using ensembles around selected points in order to
predict the local predictability of that time
series~\cite{Smith199250}.  This approach can be used to show that
different portions of a time series can exhibit varying levels of
local predictive uncertainty.  We expand on this finding later in this
paper with a time series that exhibits interesting regime shifts.

Local model-based error analysis works quite well, but it only
approximates the \emph{local} predictive uncertainty \emph{in relation
  to a fixed model}.  It cannot quantify the inherent predictability
of a time series and thus cannot be used to draw conclusions about
predictive structure that may be usable by other forecast methods.
%
%Distrubtion of error. For many methods, if your error is not normally distributed this signals that there is stil more predictve structure to that could be used by for example a larger order ARMA process. But if error is normally distributed, this suggests that you have used up all the predictive structure that **that** model can use, this doesn't quantify if preidctuce structure exists that isn't being used by this process. For example, nonlinear structure which is ignored by a linear predictor.
%
Global model-based error analysis moves in this direction.  It uses
out-of-sample error distributions, computed \emph{post facto} from a
class of models, to determine which of those models was best.  After
building an autoregressive model, for example, it is common to
calculate the errors and verify that they are normally distributed.
If they are not, that indicates that there is structure in the time
series that the model-building process was unable to capture and use.
The problem with this approach is lack of generality.  
\label{page:normal-errors}
Normally distributed errors indicate that a model has captured the
structure in the data insofar as is possible, \emph{given the
  formulation of that particular model} (viz., the best possible
linear fit to a nonlinear dataset).  This gives no indication as to
whether another modeling strategy might do better.

A practice known as deterministic/sto\-chas\-tic
modeling~\cite{Casdagli92dvsplots, weigend-book} bridges the gap
between local and global approaches to model-based error analysis.
The basic idea is to construct a series of local linear fits,
beginning with a few points and working up to to a global linear fit
that includes all known points, and then analyze how the average
out-of-sample error changes as a function of number of points in the
fit. The shape of such a graph indicates the amounts of determinism
and stochasticity are present in a time series.

@@

We build on the notion of \emph{redundancy} that was introduced on
page~\pageref{page:redundancy}, which formally quantifies how
information propagates forward through a time series:
% how much information prior observations of a system lend to future
% values
i.e., the mutual information between the past $n$ observations and the
current one.
% took this out because DCE hasn't been introduced yet: 
% 
% For example,
% this would be the amount of information that delay coordinate
% embedding captures.
The redundancy of i.i.d. random processes is zero, since all
observations are independent of one another.  On the other hand, any
deterministic system---including chaotic ones---has maximal redundancy
and thus can be perfectly predicted if observed for long
enough~\cite{weigend-book}.  In practice, it is quite difficult to
estimate the redundancy of an arbitrary, real-valued time series.
Doing so requires knowing either the Kolmolgorov-Sinai entropy or the
values of all positive Lyapunov exponents of the system.  Both of
these calculations are difficult---the latter particularly so if the
data are very noisy or the generating system is stochastic.
